% Cover Letter — Nature Communications
% Compile: pdflatex cover_letter_nature.tex

\documentclass[11pt]{letter}
\usepackage[margin=2.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\signature{Cl\'{e}ment Paulus}
\address{Independent Researcher\\clementpaulus9@gmail.com}

\begin{document}

\begin{letter}{The Editors\\Nature Communications}

\opening{Dear Editors,}

I am writing to submit the manuscript entitled \textbf{``A single
algebraic diagnostic detects structurally identical phase transitions
from quarks to the cosmic horizon''} for consideration as an Article
in \emph{Nature Communications}.

\textbf{Summary.} We report an empirical finding: the ratio of
composite integrity to fidelity (IC/$F$)---the geometric mean of
normalized measurements divided by their arithmetic mean---computed
from a single algebraic kernel, produces structurally identical
phase-transition signatures across
systems spanning 61~orders of magnitude---from quark confinement
($10^{-15}$\,m) to dark-energy--dominated cosmology ($10^{26}$\,m).
At the quark--hadron boundary, IC/$F$ drops by a factor of~23 as
color confinement drives a single measurement channel to zero;
atomic structure subsequently restores coherence 35-fold through
emergent degrees of freedom. The heterogeneity gap
$\Delta = F - \text{IC}$ spikes at every phase boundary tested,
providing a single-number, scale-independent diagnostic for
structural fragility.

\textbf{Significance.} No existing cross-scale diagnostic has been
shown to yield the same phase-transition signature at both the hadron
and cosmological scales using identical parameters and identical
equations. The mathematical guarantee behind this diagnostic---three
algebraic identities holding exactly across all 406~physical objects
tested in 20~scientific domains (15{,}014~automated proofs, zero
violations)---ensures that the cross-scale comparisons are
structurally rigorous, not correlational. The integrity bound
IC~$\leq\,F$ corresponds to the classical weighted inequality of
arithmetic and geometric means; the contribution lies not in the
bound itself but in its systematic application as a
phase-transition diagnostic across scales from subatomic to
cosmological.

\textbf{Scope.} This work is interdisciplinary by construction:
the kernel operates identically across particle physics, atomic
physics, neuroscience, cosmology, and 16~additional domains.
We believe \emph{Nature Communications} is the ideal venue because
the result bridges multiple fields through a single quantitative
framework, and the paper includes three testable predictions
amenable to independent verification by domain specialists.

\textbf{Reproducibility.} All source code, test suites,
20~domain closures, 21~versioned contracts, and raw data are
publicly available at
\url{https://github.com/calebpruett927/GENERATIVE-COLLAPSE-DYNAMICS}
under the MIT license, with permanent archival via Zenodo DOIs.
SHA-256 integrity checksums for all tracked files are maintained
and verified before every evaluation.

\textbf{Competing interests.} The author declares no competing
interests. This work was conducted independently without
institutional affiliation or external funding.

Thank you for considering this manuscript. I look forward to your
response.

\closing{Sincerely,}

\end{letter}
\end{document}
